Efficiency

### 2018-09-18 Second.Rnw
big commit

That paper, and a subsequent blog post \parencite{morrison2018a} on the topic which made the paper publically available, attracted the interest of the Senate Economics Reference Committee Inquiry into the Future of Naval Shipbuilding in Australia.  Senator Chris Ketter asked written questions on notice to the Department of Defence requesting responses to the paper specifically, and related comments from public officials, with a return date of 3rd April 2018.  I was also invited to provide evidence to the committee in a public hearing on the 7th of June 2018.  In that public hearing, I stood by the conclusion of my substantive paper, which was: ``between two otherwise identical submarines, the one with the pump jet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurance and a lower overall range than the one with the propeller."  In the presence of both the outgoing acting CEO of Naval Group Australia Brent Clark, and General Manager of Submarines (for Defence) Stephen Johnson, I pointed out that that statement was yet to be challenged, and that I'd be very surprised if anyone with technical competence or professional integrity sought to challenge it.

In their own evidence during the course of the day, neither of those two did choose to challenge it.  Senator Kim Carr requested that defence produce a written response to that report for the committee prior to their handing down of their finaly report for the Inquiry, and Defence agreed to do their best in the time available, with the head of Capability and Sustainment Group (CASG) Kim Gillis noting that ``we do hold that expertise", and Senator Carr emphasising that ``it would look pretty bloody stupid after the report's handed in''. Despite these exhortations for timeliness, in the final report of the committee, this request, and defence's acceptance of it was noted, but no acknowledgement was made of whether any response was actually received, or what it contained \parencite{SERC2018final}.

@@

@@

@@

\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{4 October 2017 - Jean-Michel Billig, Executive Director Australian submarine Program, Naval Group}]
\textit{In comments to journalists at the PACIFIC 2017 trade show, the Naval Group executive in charge of the Australian Program acknowledges that convetional propellers may be used on the Shortfin Barracuda \parencite{sbs2017}.  Subsequent reports quote Naval Group shifting responsibility for the propulsion choice to the Australian Government.}
\\
\\
(Interview transcribed from Reference \parencite{lane2017}.)

@@

\begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{25 Oct 2017 - Senate Estimates}]
\textit{Senator Hanson directly challenges Rear Admiral Gregory Sammut (then director of SEA 1000) about the pumpjet proposal.  Full exchange in \ref{SENhanson}.}
\\
\\
\textbf{Senator Hanson:}   This is why you've chosen them as the design-to go with the pump
jet. So it's not propellers, is it?

@@

\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{3 November 2017 - Department of Defence, in response to Question on Notice from Senator Xenophon from October 2017}]
\textit{As quoted in an ASPI Strategist blog post\parencite{davies2017}, and also by Senator Patrick in \ref{SENjohnson}, the Department of Defence argued in defence of the pump jet's efficiency.}
\\
\\
`A [pump jet] can provide higher propulsive efficiency across the speed range of a submarine, including the patrol speed of conventional submarines'.

@@

\begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{15 Dec 2017 - Senate Estimates}]
\textit{Senator Rex Patrick asks Stephen Johnson (General Manager Submarines) about the pumpjet proposal, and produces documents suggesting they are less efficient at low speed.  The Defence Chief Scientist, Dr Zelinsky, and Chief of Defence Force, Air Chief Marshal Binskin, and Head of Capability and Sustainment Group, Kim Gillis are also involved in the discussion, defending the choice of the pumpjet.  Senator Patrick also specifically raises the concern about the impact of the weight of the pumpjet at the extremity of the submarine.   Full exchange in \ref{SENjohnson}.}
\\
\\
\textbf{Senator PATRICK:} Mr Johnson, there's been a lot of talk about pump jets. I've now tabled a document which has come from the United States, from Connecticut. You'd be very familiar with the organisation there. That basically describes the characteristics of pump jets, their pros and cons, as any piece of engineering equipment has. The document-and there's also another document that I've taken from an ASPI site-shows the efficiency characteristics of a pump jet. Both the US document-now declassified-and the document from the ASPI website suggest that, at low speed, pump jets are relatively inefficient compared to, say, a propeller. That may well be the reason why we see a pump jet on a Mark 48 torpedo but generally not on merchant ships and so forth. Would you care to comment on that?

@@

\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{9 February 2018 - Department of Defence, in response to Question on Notice from Senator Rex Patrick in December 2017}]
\textit{Senator Patrick inquires about the nature, cost, and conclusions of earlier research into pumpjets undertaken as part of \$214 million worth of ``de-risking" studies already undertaken, as well as the schedule and cost of subsequent research and development of the pumpjet.  Prior research consisted of a single 99 page report obtained via literature review and discussion with experts at a cost of \$68,825.  Ongoing research included \$1.1million spent to date to develop a generic submarine and pumpjet design, compuational and analytic models, and validation of these models through experimental studies.  The development and test program is not scheduled to commence until the last quarter of 2018.  Full questions and responses can be found in \ref{QON148}, \ref{QON150}, \ref{QON151}}
\\
\\
``The study concluded that pumpjets have the potential to deliver increased propulsive efficiency and lower shaft torque requirements, relative to an open propeller. This should lead to beneficial platform systems and operational impacts to increase vessel speed and/or submerged endurance.
The study developed a parametric model of pumpjet thrust and efficiency that has been used as the basis for more recent studies into pumpjet performance."
\\
\\
``Specifications for the pump jet are being developed as design of the Future Submarine proceeds. Initial inputs will be developed in the last quarter of 2018, which marks the beginning of the development and test program for the pump jet. Development, testing, and assurance will be conducted by several parties including Naval Group, Direction generale de l'armement (DGA), and DSTG. Testing and assurance activities will take place in France (including at the premises of the DGA Techniques Hydrodynamiques in Val de Reuil) and Australia. Testing will be largely completed by CDR."

@@

\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{3 April 2018 - Department of Defence, in response to Question on Notice from Senator Chris Ketter in December 2018}]
\textit{Senator Ketter asks Question 1, referencing specifically Greg Sammut in \parencite{sammut2017} and my paper \parencite{morrison2018} asks specifically whether Defence maintians that the pumpjet can be efficient across the entire speed range. Full question can be found in \ref{QON1}.}
\\
\\
Question: Does defence still maintain that a pump-jet can be efficient across the entire speed range?
\\
\\
Answer: ``The efficiency of pump-jets designed for submarine use cannot be determined through comparisons with water-jets and ducted propellers, which are different in their operation from pump-jets. Moreover, the global efficiency of submarine pump-jets cannot be derived from `open water' tests of pump-jets, where the pump-jet is assessed independently of the hull of the submarine. A correctly designed pump-jet, suited to the aft hull shape of the submarine, achieves better optimisation of efficiency and stealth than any propeller, even at low speeds."

@@

Overall, this response casts serious doubt on the type of communication and advice which has flowed between Naval Group and the Department of Defence regarding the pumpjet to date.  As a ship-builder which had designed and manufactured both pumpjets and propellers, and fitted them to both diesel-electric as well as nuclear submarines, there could be no excuse for Naval Group to not have a well-informed opinion on this matter.  For Defence not to have obtained that opinion from DCNS at the time of considering the bid would be a matter of incomprehensible oversight.  It is made all the more difficult to believe given the scrutiny applied to the matter in the Senate in 2012 \ref{SENmoffit}, and the program director's assurance that the work was being done to assess its suitability.

@@

It is possible to believe that Defence's own embarrassment at having been unable to detect the insincerity in Naval Group's offer of the pumpjet, and inability to reach their own technical conclusions quickly enough to detect the fraud, may motivate the present actions.  This amount to a disasterous failure on the part of the public service to act with integrity and in the national interest.  However, if Defence had truely been un-wittingly deceived, it is difficult to see how they could not recover their pride (and regain a necessary measure of respect from their build partner) by quickly adopting a more combative stance, calling out the deceptions on the part contractors part, and quickly exacting whatever contractual reparations they might be entitled to.  If the contractual reparations were not sufficiently strong, the Commonwealth's freedom to not proceed to a build contract (from the initial design contract) or re-introduce competition into the design process as many have suggested should occur in any case \parencite{stanford2017}, would have provided ample commercial leverage to reset the relationship in a manner which was better aligned with the Commonwealth's interest.

One must at this point give consideration to the possibly that Defence had not been entirely unwittingly deceived, and was actually complicit in their own deception, and it's present upkeep. This would mean that the Department of Defence and the Navy, or at least a few of very senior people within those organisations, had to conspired allow a proposal which was known to be inappropriate to the Commonwealth's needs to appear victorious in the CEP,  presumeably in order to favour a bidder for whom no more defensible explanation could be found.  It appears unlikely that there could be any decent motive for such a course of action.  However, one must evaluate the probabilities of the necessary oversights for such an eventuatlity occurring by accident or by arrangement.  If the pumpjet was known by DCNS to be an unsuitable proposition for a diesel-electric submarine, a seemingly unlikely degree of naive incompetence would be required for Defence to fail to ascertain as much during the Competitive Evaluation Process.  To remain genuinely unwitting for two subsequent years of design development would seem even more so.  The possibility of some corruption influencing the selection of Australia's \$50bn future submarine contract must at least be considered.

\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{3 April 2018 - Department of Defence, in response to Question on Notice from Senator Chris Ketter in December 2018}]
\textit{Senator Ketter asks Question 3, which focuses specifically on whether Defence believes the pumpjet is acoustically superior across the entire speed range, or just at higher speeds, referencing my paper \parencite{morrison2018}, Herve Guillou \parencite{guillou2017} and Costello and Autret \parencite{autret2016}.  Full question and response can be found in \ref{QON3}.}
\\
\\
Question: Is Defence of the belief that the pump-jet is quieter than propellers at all operating speeds, or is it the case, as suggested by Mr Morrison, that at some low speeds propellers are acoustically superior, and only at higher speeds is the pump-jet acoustically superior?
\\
\\
Answer: ``See response to Senate Economics Reference Committee Question on Notice, Q1 – Pump Jet Efficiency Across the Entire Speed Range."
\end{tcolorbox}
The response to Q1 - a question specifically and only asking a question about \textit{efficiency} of the pumpjet, avoids contradicting the evidence directly by conflating the question with acoustics.  To avoid the answering a question that is specifically about acoustics by referring to that same answer which incorporates efficiency is nothing more than a point-blank refusal to answer either question directly.

\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{17 May 2018 - Brent Clark, Acting CEO Naval Group Australia}]
\textit{In an interview that was published in a post\parencite{guillou2017} on the ASPI Strategist the Autralian CEO of Naval group explains to Brendan Nicholson why Naval Group is recommending the pumpjet for the Shortfin Barracuda, and answers some critics.}
\\
\\
The company building the Navy's 12 new submarines says they'll be the world's first conventional boats big enough to effectively use the company's pump jet propulsion system. French Naval Group Australia says that its propulsor will prove much quieter than a propeller on the 4,700‑tonne submarine it has dubbed the 'Shortfin Barracuda'.
\\
\\
Naval Group Australia's interim CEO, Brent Clark, tells The Strategist that the decision about the best propulsion system for the submarine will be made by the Defence Department, but that he hadn't had any indication that it would chose anything but a propulsor. 'We've had every indication that they like what was offered.'
\\
\\
The propulsor that Naval Group is recommending for its Shortfin Barracuda consists of a propeller within a duct to reduce noise and the water disturbance known as cavitation, which can give away a boat's position to hunters. Clark says he's read many negative comments about propulsors, but some critics used very old studies to try and justify their argument.
\\
\\
"A lot of people don't fully comprehend, I think, that yes, a propulsor on a 1,500-tonne coastal submarine would make no sense. But we're not building a 1,500-tonne coastal submarine. We're building quite a large submarine. And this is the first time that a conventional submarine's been large enough to take a propulsor."

@@

This is also the first time that the argument that the size of the submarine is a crucial enabling difference that makes the pumpjet viable is made.  It is the closest that Naval Group has come from making a firm statement of faith in the suitability of the technology for the Australian diesel-elctric submarine. As I discussed in another blog post in ASPI Strategist \parencite{morrison2018b} shortly afterwards, this is a remarkeably concession that is consistent with the physics that I outlined, since the inefficiency of the pumpjet would likely drive demand for additional fuel and batteries (hence size) to meet requirements, which is more of a concession of the detrimental impact of the pumpjet on the submarine.  This particular blog post was acknowledged on twitter by one former US Navy submariner and submarine enthusiast as an ``Awesome article about pump jet efficiencies and viability".  Naval Group was tagged in the re-tweet, which was subsequently re-tweeted and liked by naval experts, publications, commentators and authors from around the world.

\begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{07 Jun 2018 - Senate Economic References Committee}]
\textit{The author presents evidence to the final public hearing of this committee's Inquiry into the Future of Australia's Naval Shipbuilding Industry, challenging the pumpjet choice.  Brent Clark, the outgoing CEO Naval Group Australia, Stephen Johnson, the General Manager Submarines from Defence, as well as Kim Gillis, Head of Capability and Sustainment Group at Defence, are all invited to respond to author's evidence.  Full extracts of all discussion on the pumpjet topic are in  \ref{SENmorrison}.}
\\
\\
\textbf{Mr Morrison:} ...In my report, I made the conclusive comparison that, between two otherwise identical submarines, the one with the pump jet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurance and a lower overall range than the one with the propeller. This statement, so far as I know, has yet to be challenged, and \textbf{I'd be very surprised if anyone with technical competence or professional integrity sought to challenge it.}
...

@@

Naval Group here steadfastly refuses to offer a clear opinion on the pumpjet. Only months earlier Defence was asked a specific question on notice about whether they thought Naval Group maintained that a pumpjet was suitable for a diesel-electric submarine.  Clarke's interview on the Strategist, which seemed to be strongest statement of faith from Naval group in the suitability of the pumpjet for a conventional submarine (perhaps in an effort to patch up the noteable refusal to answer the question) had been answered publically by the same person who had just appeared and he was being asked to respond to.  It might have been expected that Mr Clark would at least attempt to demonstrate conviction about the merits of the proposal.  Instead, he described the information that they were supplying as a classified, government-to-government exchange, as if he were an impartial bystander.  No assurance of the proposals superiority, suitability, or viability in any respect was made.

@@

\textbf{Senator REYNOLDS:} Mr Gillis!
\\
\\
...
\\
\\
\textbf{Senator KIM CARR:} I contend that we don't have the technical expertise to evaluate the report. It may well be that we recommend that further work be done on it, but I'd want to see your response to this-to give us a critique of it; your perspective as to the report. He's made some assertions to the committee. You no doubt will get the Hansard pretty quickly, if the secretary can make sure that that's available to the officers.

@@

This exchange demonstrates a dismissive attitude towards the evidence I had presented, which borders on contempt. The attitude is clearly pervasive and supported in the higest levels of the defence department.  The General Manager Submarines, Stephen Johnson, claims he simply cannot recall what he said about pumpjets being related to pods on cruise ships, despite being physically present in the same room when I pointedly quoted and contradicted him.  This being pointed out, he claimed to simply not to be paying attention, which won the support of the Head of Capapbility and Sustainment Group, Kim Gillis.  Despite the obvious shock and exclamation of the senators, Mr Gillis went on to attempt to dismiss my evidence as just another opinion which could probably to be dismissed along with 99\% of the others.

This level of ambivalence towards external criticism from a such a senior official on such a topic as this should be alarming.  The topic of the pumpjet had such sustained criticism and interest from various quarters, including multiple questions from senators from all sides of government over six years, reports in the ABC, ASPI Strategist, Defence Connect, Insight Economics, and many others.  My paper was commissioned by a concerned citizen who wanted to have a better investigation into the physics underlying the controversy which had already received considerable attention, albeit with confusing results.

Such an attitude raises questions about whether Mr Gillis's evidence is even given honestly.  For instance, Mr Gillis' claim that he had not read the report, since ``we weren't actually  given a copy" warrants serious attention.  My paper was made publically available with the blog-post summarising it on the 14th of February, and as well as in my reply to Brend Clark on  the 24th of May.  Moreover, his department had received written Questions on Notice from Senator Chris Ketter, acting on behalf of this very committee, specifically requesting written responses to my paper with a return date of 3 April.  His department did indeed responded to prior to the hearing, as noted above with responses which blatantly refused to answer the questions. Furthermore this particular hearing was also especially long in the making, since it was once postponed on account of Defence refusing to make officials available to attend.  My presence as a witness was requested in February when the proposed date was still in May, and the list of witnesses was provided publically circulated during the weeks prior to the hearing.

In such circumstances, claims to be ignorant of my report, and an attempt to lump it with the other 99\% of dissenting correspondence, seem about as reasonable as Mr Johnson's claim that he simply wasn't paying attention when he was being quoted and refuted before the committee earlier in the day.  That this is the type of responses being given by the most senior officials who are responsible for the largest defence procurement in Australia's history should be alarming.

>>>>

That paper, and a subsequent blog post \parencite{morrison2018a} on the topic which made the paper publically available, attracted the interest of the Senate Economics Reference Committee Inquiry into the Future of Naval Shipbuilding in Australia.  Senator Chris Ketter asked written questions on notice to the Department of Defence requesting responses to the paper specifically, and related comments from public officials, with a return date of 3rd April 2018.  I was also invited to provide evidence to the committee in a public hearing on the 7th of June 2018.  In that public hearing \ref{SENmorrison}, I stood by the conclusion of my substantive paper, which was: ``between two otherwise identical submarines, the one with the pump jet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurance and a lower overall range than the one with the propeller."  In the presence of both the outgoing acting CEO of Naval Group Australia Brent Clark, and General Manager of Submarines (for Defence) Stephen Johnson, I pointed out that that statement was yet to be challenged, and that I'd be very surprised if anyone with technical competence or professional integrity sought to challenge it.

In their own evidence during the course of the day, neither of those two did choose to challenge it.  Senator Kim Carr requested that defence produce a written response to that report for the committee prior to their handing down of their finaly report for the Inquiry, and Defence agreed to do their best in the time available, with the head of Capability and Sustainment Group (CASG) Kim Gillis assuring the committee that ``We  will  provide  the  committee  with  the  best  possible  advice  wecan  in  the  shortest  possible  time  frame  to  meet  your  requirements  so  that  you  can  be  betterinformed, because we do hold that expertise." Senator Carr emphasised the importance of timeliness that ``it would look pretty bloody stupid after the report's handed in''. Despite these exhortations, in the final report of the committee this request, and defence's acceptance of it was noted, but no acknowledgement was made of whether any response was actually received, or what it contained \parencite{SERC2018final}.

@@

\subsection{RADM Rowan Moffit, Head of SEA 1000, before the Senate}

@@

\subsection{Naval Group on ASPI Strategist and own website}

@@

\subsection{Naval Group at PACIFIC 2017 and in ASPI Strategist}
\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{4 October 2017 - Jean-Michel Billig, Executive Director Australian submarine Program, Naval Group}]
\textit{In comments to journalists at the PACIFIC 2017 trade show, the Naval Group executive in charge of the Australian Program acknowledges that convetional propellers may be used on the Shortfin Barracuda \parencite{sbs2017}.  Subsequent reports quote Naval Group shifting responsibility for the propulsion choice to the Australian Government.}
\\
\\
(Interview transcribed from Reference \parencite{lane2017}.)

@@

\subsection{RADM Sammut, Head of SEA 1000, to Senate and ASPI Strategist}
\begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{25 Oct 2017 - Senate Estimates}]
\textit{Senator Hanson directly challenges Rear Admiral Gregory Sammut (then director of SEA 1000) about the pumpjet proposal.  Full exchange in \ref{SENhanson}.}
\\
\\
\textbf{Senator Hanson:}   This is why you've chosen them as the design-to go with the pump
jet. So it's not propellers, is it?

@@

\subsection{Department of Defence, to Senator Xenophon's Question on Notice}
\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{3 November 2017 - Department of Defence, in response to Question on Notice from Senator Xenophon from October 2017}]
\textit{As quoted in an ASPI Strategist blog post\parencite{davies2017}, and also by Senator Patrick in \ref{SENjohnson}, the Department of Defence argued in defence of the pump jet's efficiency.}
\\
\\
`A [pump jet] can provide higher propulsive efficiency across the speed range of a submarine, including the patrol speed of conventional submarines'.

@@

\subsection{Stephen Johnson, General Manager Submarines, before the Senate}
\begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{15 Dec 2017 - Senate Estimates}]
\textit{Senator Rex Patrick asks Stephen Johnson (General Manager Submarines) about the pumpjet proposal, and produces documents suggesting they are less efficient at low speed.  The Defence Chief Scientist, Dr Zelinsky, and Chief of Defence Force, Air Chief Marshal Binskin, and Head of Capability and Sustainment Group, Kim Gillis are also involved in the discussion, defending the choice of the pumpjet.  Senator Patrick also specifically raises the concern about the impact of the weight of the pumpjet at the extremity of the submarine.   Full exchange in \ref{SENjohnson}.}
\\
\\
\textbf{Senator PATRICK:} Mr Johnson, there's been a lot of talk about pump jets. I've now tabled a document which has come from the United States, from Connecticut. You'd be very familiar with the organisation there. That basically describes the characteristics of pump jets, their pros and cons, as any piece of engineering equipment has. The document-and there's also another document that I've taken from an ASPI site-shows the efficiency characteristics of a pump jet. Both the US document-now declassified-and the document from the ASPI website suggest that, at low speed, pump jets are relatively inefficient compared to, say, a propeller. That may well be the reason why we see a pump jet on a Mark 48 torpedo but generally not on merchant ships and so forth. Would you care to comment on that?

@@

\subsection{Department of Defence, to Senator Patrick's Questions on Notice}
\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{9 February 2018 - Department of Defence, in response to Question on Notice from Senator Rex Patrick in December 2017}]
\textit{Senator Patrick inquires about the nature, cost, and conclusions of earlier research into pumpjets undertaken as part of \$214 million worth of ``de-risking" studies already undertaken, as well as the schedule and cost of subsequent research and development of the pumpjet.  Prior research consisted of a single 99 page report obtained via literature review and discussion with experts at a cost of \$68,825.  Ongoing research included \$1.1million spent to date to develop a generic submarine and pumpjet design, compuational and analytic models, and validation of these models through experimental studies.  The development and test program is not scheduled to commence until the last quarter of 2018.  Full questions and responses can be found in \ref{QON148}, \ref{QON150}, \ref{QON151}}
\\
\\
``The study concluded that pumpjets have the potential to deliver increased propulsive efficiency and lower shaft torque requirements, relative to an open propeller. This should lead to beneficial platform systems and operational impacts to increase vessel speed and/or submerged endurance.
The study developed a parametric model of pumpjet thrust and efficiency that has been used as the basis for more recent studies into pumpjet performance."
\\
\\
``Specifications for the pump jet are being developed as design of the Future Submarine proceeds. Initial inputs will be developed in the last quarter of 2018, which marks the beginning of the development and test program for the pump jet. Development, testing, and assurance will be conducted by several parties including Naval Group, Direction generale de l'armement (DGA), and DSTG. Testing and assurance activities will take place in France (including at the premises of the DGA Techniques Hydrodynamiques in Val de Reuil) and Australia. Testing will be largely completed by CDR."

@@

\subsection{Department of Defence, to Senator Ketter's Questions on Notice}
\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{3 April 2018 - Department of Defence, in response to Question on Notice from Senator Chris Ketter in December 2018}]
\textit{Senator Ketter asks Question 1, referencing specifically Greg Sammut in \parencite{sammut2017} and my paper \parencite{morrison2018} asks specifically whether Defence maintians that the pumpjet can be efficient across the entire speed range. Full question can be found in \ref{QON1}.}
\\
\\
Question: Does defence still maintain that a pump-jet can be efficient across the entire speed range?
\\
\\
Answer: ``The efficiency of pump-jets designed for submarine use cannot be determined through comparisons with water-jets and ducted propellers, which are different in their operation from pump-jets. Moreover, the global efficiency of submarine pump-jets cannot be derived from `open water' tests of pump-jets, where the pump-jet is assessed independently of the hull of the submarine. A correctly designed pump-jet, suited to the aft hull shape of the submarine, achieves better optimisation of efficiency and stealth than any propeller, even at low speeds."

@@

Overall, this response casts serious doubt on the type of communication and advice which has flowed between Naval Group and the Department of Defence regarding the pumpjet to date.  As a ship-builder which had designed and manufactured both pumpjets and propellers, and fitted them to both diesel-electric as well as nuclear submarines, there could be no excuse for Naval Group to not have a well-informed opinion on this matter.  For Defence not to have obtained that opinion from DCNS at the time of considering the bid would be a matter of incomprehensible oversight.  It is made all the more difficult to believe given the scrutiny applied to the matter in the Senate in 2012 \ref{SENMoffit}, and the program director's assurance that the work was being done to assess its suitability.

@@

It is possible to believe that Defence's own embarrassment at having been unable to detect the insincerity in Naval Group's offer of the pumpjet, and inability to reach their own technical conclusions quickly enough to detect the fraud, may motivate the present actions.  This would  amount to a disasterous failure on the part of the public service to act with integrity and in the national interest.  However, if Defence had truely been un-wittingly deceived, it is difficult to see how they could not recover their pride (and regain a necessary measure of respect from their design partner) by quickly adopting a more combative stance, calling out the deceptions on the part contractors part, and quickly exacting whatever contractual reparations they might be entitled to.  If the contractual reparations were not sufficiently strong, the Commonwealth's freedom to not proceed to a build contract (from the initial design contract) or re-introduce competition into the design process as many have suggested should occur in any case \parencite{stanford2017}, would have provided ample commercial leverage to reset the relationship in a manner which was better aligned with the Commonwealth's interest.

One must at this point give consideration to the possibly that Defence had not been entirely unwittingly deceived, and was actually complicit in their own deception, including its present upkeep. This would mean that the Department of Defence and the Navy, or at least a few of very senior people within those organisations, had conspired allow a proposal which was known to be inappropriate to the Commonwealth's needs to appear victorious in the CEP,  presumeably in order to favour a bidder for whom no more defensible explanation could be found.  It appears unlikely that there could be any decent motive for such a course of action, and would lead invariably to the grave conclusion that some kind of corruption has occurred.  In the circumstances, one must evaluate the probabilities of the necessary oversights for such an eventuatlity occurring by accident against the occurring by arrangement.  If the pumpjet was known by DCNS to be an unsuitable proposition for a diesel-electric submarine, a seemingly unlikely degree of naive incompetence would be required for Defence to fail to ascertain as much during the Competitive Evaluation Process.  To remain genuinely unwitting for two subsequent years of design development would seem even more so.  The possibility of some corruption influencing the selection of Australia's \$50bn future submarine contract must at least be considered.

\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{3 April 2018 - Department of Defence, in response to Question on Notice from Senator Chris Ketter in December 2018}]
\textit{Senator Ketter asks Question 3, which focuses specifically on whether Defence believes the pumpjet is acoustically superior across the entire speed range, or just at higher speeds, referencing my paper \parencite{morrison2018}, Herve Guillou \parencite{guillou2017} and Costello and Autret \parencite{autret2016}.  Full question and response can be found in \ref{QON3}.}
\\
\\
Question: Is Defence of the belief that the pump-jet is quieter than propellers at all operating speeds, or is it the case, as suggested by Mr Morrison, that at some low speeds propellers are acoustically superior, and only at higher speeds is the pump-jet acoustically superior?
\\
\\
Answer: ``See response to Senate Economics Reference Committee Question on Notice, Q1 - Pump Jet Efficiency Across the Entire Speed Range."
\end{tcolorbox}
The response to Q1 - a question specifically and only asking a question about \textit{efficiency} of the pumpjet, avoids contradicting the evidence directly by conflating the question with acoustics.  To avoid the answering a question that is specifically about acoustics by referring to that same answer which incorporates efficiency amounts to nothing more than a point-blank refusal to answer either question directly.

\subsection{Brent Clark, Acting CEO Naval Group Australia, in ASPI Strategist}
\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{17 May 2018 - Brent Clark, Acting CEO Naval Group Australia}]
\textit{In an interview that was published in a post\parencite{guillou2017} on the ASPI Strategist the Autralian CEO of Naval group explains to Brendan Nicholson why Naval Group is recommending the pumpjet for the Shortfin Barracuda, and answers some critics.}
\\
\\
The company building the Navy's 12 new submarines says they'll be the world's first conventional boats big enough to effectively use the company's pump jet propulsion system. French Naval Group Australia says that its propulsor will prove much quieter than a propeller on the 4,700-tonne submarine it has dubbed the `Shortfin Barracuda'.
\\
\\
Naval Group Australia's interim CEO, Brent Clark, tells The Strategist that the decision about the best propulsion system for the submarine will be made by the Defence Department, but that he hadn't had any indication that it would chose anything but a propulsor. `We've had every indication that they like what was offered.'
\\
\\
The propulsor that Naval Group is recommending for its Shortfin Barracuda consists of a propeller within a duct to reduce noise and the water disturbance known as cavitation, which can give away a boat's position to hunters. Clark says he's read many negative comments about propulsors, but some critics used very old studies to try and justify their argument.
\\
\\
``A lot of people don't fully comprehend, I think, that yes, a propulsor on a 1,500-tonne coastal submarine would make no sense. But we're not building a 1,500-tonne coastal submarine. We're building quite a large submarine. And this is the first time that a conventional submarine's been large enough to take a propulsor."

@@

This is also the first appearance of the argument that the size of the submarine is a crucial factor that makes the pumpjet viable.  It is the closest that Naval Group has come to making a firm statement of faith in the suitability of the technology for the Australian diesel-elctric submarine. As I discussed in another blog post in ASPI Strategist \parencite{morrison2018b} shortly afterwards, this is a remarkeably concession that is consistent with the physics that I outlined, since the inefficiency of the pumpjet would likely drive demand for additional fuel and batteries (hence size) to meet requirements.  As such, this admission is more of a concession of the detrimental impact of the pumpjet on the submarine.  This particular blog post was acknowledged on twitter by one former US Navy submariner and submarine enthusiast as an ``Awesome article about pump jet efficiencies and viability".  Naval Group was tagged in the re-tweet, which was subsequently re-tweeted and liked by naval experts, publications, commentators and authors from around the world.

\subsection{Morrison Senate evidence and Clark, Johnson, Gillis responses}
\begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{07 Jun 2018 - Senate Economic References Committee}]
\textit{The author presents evidence to the final public hearing of this committee's Inquiry into the Future of Australia's Naval Shipbuilding Industry, challenging the pumpjet choice.  Brent Clark, the outgoing CEO Naval Group Australia, Stephen Johnson, the General Manager Submarines from Defence, as well as Kim Gillis, Head of Capability and Sustainment Group at Defence, are all invited to respond to author's evidence.  Full extracts of all discussion on the pumpjet topic are in  \ref{SENmorrison}.}
\\
\\
\textbf{Mr Morrison:} ...In my report, I made the conclusive comparison that, between two otherwise identical submarines, the one with the pump jet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurance and a lower overall range than the one with the propeller. This statement, so far as I know, has yet to be challenged, and \textbf{I'd be very surprised if anyone with technical competence or professional integrity sought to challenge it.}
...

@@

Naval Group here steadfastly refuses to offer a clear opinion on the pumpjet. Only months earlier Defence was asked a specific question on notice about whether they thought Naval Group maintained the opinion that a pumpjet was suitable for a diesel-electric submarine.  Clarke's interview on the Strategist, which seemed to be strongest statement of faith from Naval group in the suitability of the pumpjet for a conventional submarine (perhaps in an effort to patch up the noteable refusal to answer the question) had been answered publically by the same person who had just appeared and he was being asked to respond to.  It might have been expected that Mr Clark would at least attempt to demonstrate conviction about the merits of the proposal.  Instead, he described the information that they were supplying as a classified, government-to-government exchange, as if he were an impartial bystander.  No assurance of the proposals superiority, suitability, or viability in any respect was made.

@@

...
\end{tcolorbox}

@@

This exchange demonstrates a dismissive attitude towards the evidence I had presented, which borders on contempt. The attitude is clearly pervasive and supported in the higest levels of the defence department.  The General Manager Submarines, Stephen Johnson, claims he simply cannot recall what he said about pumpjets being related to pods on cruise ships, despite being physically present in the same room when I pointedly quoted and contradicted him.  This being pointed out, he claimed to simply not to be paying attention, which won the support of the Head of Capapbility and Sustainment Group, Kim Gillis.  Despite the obvious shock and exclamation of the senators, Mr Gillis went on to attempt to dismiss my evidence as just another opinion which could probably to be dismissed along with 99\% of the others.  This is a sustained and deliberate effort to adopt a dismissive attitude in the face of serious criticism.

This level of ambivalence towards external criticism from a such a senior official on such a topic as this is worth noting.  The topic of the pumpjet had such sustained criticism and interest from various quarters, including multiple questions from senators from all sides of government over six years, reports in the ABC, ASPI Strategist, Defence Connect, Insight Economics, and many others.  My paper was commissioned by a concerned citizen who wanted to have a better investigation into the physics underlying the controversy which had already received considerable attention, albeit with confusing results.

Such an openly dismissive attitude, exemplified by Mr Gillis' ``not all of this is rivetting" raises questions about whether Mr Gillis's evidence is even given genuinely and honestly.  For instance, Mr Gillis' claim that he had not read the report, since ``we weren't actually  given a copy" warrants serious attention.  My paper was made publically available with the blog-post summarising it on the 14th of February, and as well as in my reply to Brend Clark on  the 24th of May.  Moreover, his department had received written Questions on Notice from Senator Chris Ketter, acting on behalf of this very committee, specifically requesting written responses to my paper with a return date of 3 April.  His department most probably responded to prior to the hearing, as noted above, with responses which blatantly refused to answer the questions. \footnote{However, I was not made aware either of the questions or their answers prior to hearing.} Furthermore this particular hearing was also especially long in the making, since it was once postponed on account of Defence refusing to make officials available to attend.\parencite{adm2018}  My presence as a witness was requested in February when the proposed date was still in May, and the list of witnesses was publically circulated during the weeks prior to the hearing.  In such circumstances, claims to be essentially ignorant of my report and pretend it could probably be dismissed after the reading of the first page, should be examined critically.
